The Unicode Cookbook \linebreak for Linguists

### 2018-03-12 book/chapters/pitfalls.tex
fixes section reference
Resolves issue #29

Third, a single glyph may be a combination of multiple characters. For example, 
the ligature <ﬁ>, a single glyph, is the result of two
characters, <f> and <i>, that have undergone glyph substitution by font
rendering (see also Section~\ref{pitfall-faulty-rendering}). Like
contextually-determined glyphs, ligatures are (intended) artefacts of text
processing instructions. Finally, a single glyph may be a part of a
character, as exemplified by diacritics like the diaeresis <\dia{0308}> in <ë>.

@@

Inspecting the Unicode Standard, there appear to be special Unicode characters
to glue together characters into larger tailored grapheme clusters,
specifically the \textsc{zero width joiner} at \uni{200D} and the
\textsc{combining grapheme joiner} at \uni{034F}. However, these characters are
confusingly named (cf.~Section~\ref{pitfall-names}). Both codepoints actually do
not join characters, but explicitly separate them. The zero-width joiner (ZWJ)
can be used to solve special problems related to ordering (called \textit{collation}
in Unicode parlance). The combining grapheme joiner (CGJ) can be used to
separate characters that are not supposed to form ligatures.

@@

Further, we suggest to always install at least one so-called \textsc{fall-back
font}, which provides glyphs that at least show the user some information about
the underlying encoded character. Apple Macintoshes have such a font (which is
invisible to the user), which is designed by Michael Everson and made available
for other systems through the Unicode Consortium.\footnote{The Apple/Everson
fallback font is available for non-Macintosh users at \newline
\url{http://www.unicode.org/policies/lastresortfont\_eula.html}.} Further, the
\textsc{GNU Unifont} is a clever way to produce bitmaps approximating the
intended glyph of each available character, updated to Unicode 7.0.\footnote{The
GNU Unifont is available at \url{http://unifoundry.com/unifont.html}.} Finally,
the Summer Institute of Linguistics provides a \textsc{SIL Unicode BMP Fallback
Font}, currently available up to Unicode version 6.1. This font does not even
attempt to show a real glyph, but only shows the hexadecimal code inside a box
for each character, so a user can at least see the Unicode codepoint of the
character to be displayed.\footnote{The SIL Unicode BMP Fallback Font is
available at \newline \url{http://scripts.sil.org/UnicodeBMPFallbackFont}.}

@@

A similar complaint to missing glyphs, discussed previously, is that while there
might be a glyph being displayed, it does not look right. There are two
reasons for unexpected visual display, namely automatic font substitution and
faulty rendering. Like missing glyphs, any such problems are independent from
the Unicode Standard. The Unicode Standard only includes very general
information about characters and leaves the specific visual display to others to
decide on. Any faulty display is thus not to be blamed on the Unicode
Consortium, but on a complex interplay of different mechanisms happening in a
computer to turn Unicode codepoints into visual symbols. We will only sketch a
few aspects of this complex interplay here.

@@

The actual intended meaning of a Unicode codepoint is a combination of the
name, the block and the character properties (see
Chapter~\ref{the-unicode-approach}). Further details about the underlying intentions 
with which a character should be used
are only accessible by perusing the actual decisions of the Unicode Consortium.
All proposals, discussions and decisions of the Unicode Consortium are publicly
available. Unfortunately there is not (yet) any way to easily find everything
that is ever proposed, discussed and decided in relation to a specific
codepoint of interest, so many of the details are often somewhat
hidden.\footnote{All proposals and other documents that are the basis of Unicode
decisions are available at \url{http://www.unicode.org/L2/all-docs.html}. The
actual decisions that make up the Unicode Standard are documented in the minutes
of the Unicode Technical Committee, available at
\url{http://www.unicode.org/consortium/utc-minutes.html}.}

@@

% ==========================
\section{Pitfall: Encodings}
\label{encoding}
% ==========================

@@

\begin{itemize}
   \item To prevent strange boxes instead of nice glyphs, always install a few
         fonts with a large glyph collection and at least one fall-back font
         (see Section~\ref{pitfall-missing-glyphs}).
   \item Unexpected visual impressions of symbols does not necessarily mean that
         the actual encoding is wrong. It is mostly a problem of faulty
         rendering (see Section~\ref{pitfall-faulty-rendering}).
   \item Do not trust the names of codepoints as a definition of the character
         (see Section~\ref{pitfall-names}). Also do not trust Unicode blocks as
         a strategy to find specific characters (see
         Section~\ref{pitfall-blocks})
   \item To ensure consistent encoding of texts, apply Unicode normalization
         (NFC or NFD, see Section~\ref{pitfall-canonical-equivalence}).
   \item To prevent remaining inconsistencies after normalization, for example 
         stemming from homoglyphs (see Section~\ref{pitfall-homoglyphs}) 
         or from missing canonical equivalence in the Unicode Standard
         (see Section~\ref{pitfall-absence-of-equivalence})
         use orthography profiles (see Chapter~\ref{orthography-profiles}).
   \item To deal with tailored grapheme clusters
         (Section~\ref{pitfall-characters-are-not-graphemes}), use Unicode Locale 
         Descriptions, or orthography profiles 
         (see Chapter~\ref{orthography-profiles}).
   \item As a preferred file format, use Unicode Format UTF-8 in 
         Normalization Form Composition (NFC) with LF line endings, 
         but without byte order mark (BOM), whenever possible (see 
         Section~\ref{pitfall-file-formats}). This last nicely cryptic 
         recommendation has T-shirt potential:

>>>>

Third, a single glyph may be a combination of multiple characters. For example, 
the ligature <ﬁ>, a single glyph, is the result of two
characters, <f> and <i>, that have undergone glyph substitution by font
rendering (see also Section~\ref{pitfall-faulty-rendering}). Like
contextually-determined glyphs, ligatures are (intended) artifacts of text
processing instructions. Finally, a single glyph may be a part of a
character, as exemplified by diacritics like the diaeresis <\dia{0308}> in <ë>.

@@

Inspecting the Unicode Standard, there appear to be special Unicode characters
to glue together characters into larger tailored grapheme clusters,
specifically the \textsc{zero width joiner} at \uni{200D} and the
\textsc{combining grapheme joiner} at \uni{034F}. However, these characters are
confusingly named (cf.~Section~\ref{pitfall-names}). Both code points actually do
not join characters, but explicitly separate them. The zero-width joiner (ZWJ)
can be used to solve special problems related to ordering (called \textit{collation}
in Unicode parlance). The combining grapheme joiner (CGJ) can be used to
separate characters that are not supposed to form ligatures.

@@

Further, we suggest to always install at least one so-called \textsc{fall-back
font}, which provides glyphs that at least show the user some information about
the underlying encoded character. Apple Macintoshes have such a font (which is
invisible to the user), which is designed by Michael Everson and made available
for other systems through the Unicode Consortium.\footnote{The Apple/Everson
fallback font is available for non-Macintosh users at \newline
\url{http://www.unicode.org/policies/lastresortfont\_eula.html}.} Further, the
\textsc{GNU Unifont} is a clever way to produce bitmaps approximating the
intended glyph of each available character, updated to Unicode 7.0.\footnote{The
GNU Unifont is available at \url{http://unifoundry.com/unifont.html}.} Finally,
the Summer Institute of Linguistics provides a \textsc{SIL Unicode BMP Fallback
Font}, currently available up to Unicode version 6.1. This font does not even
attempt to show a real glyph, but only shows the hexadecimal code inside a box
for each character, so a user can at least see the Unicode code point of the
character to be displayed.\footnote{The SIL Unicode BMP Fallback Font is
available at \newline \url{http://scripts.sil.org/UnicodeBMPFallbackFont}.}

@@

A similar complaint to missing glyphs, discussed previously, is that while there
might be a glyph being displayed, it does not look right. There are two
reasons for unexpected visual display, namely automatic font substitution and
faulty rendering. Like missing glyphs, any such problems are independent from
the Unicode Standard. The Unicode Standard only includes very general
information about characters and leaves the specific visual display to others to
decide on. Any faulty display is thus not to be blamed on the Unicode
Consortium, but on a complex interplay of different mechanisms happening in a
computer to turn Unicode code points into visual symbols. We will only sketch a
few aspects of this complex interplay here.

@@

The actual intended meaning of a Unicode code point is a combination of the
name, the block and the character properties (see
Chapter~\ref{the-unicode-approach}). Further details about the underlying intentions 
with which a character should be used
are only accessible by perusing the actual decisions of the Unicode Consortium.
All proposals, discussions and decisions of the Unicode Consortium are publicly
available. Unfortunately there is not (yet) any way to easily find everything
that is ever proposed, discussed and decided in relation to a specific
code point of interest, so many of the details are often somewhat
hidden.\footnote{All proposals and other documents that are the basis of Unicode
decisions are available at \url{http://www.unicode.org/L2/all-docs.html}. The
actual decisions that make up the Unicode Standard are documented in the minutes
of the Unicode Technical Committee, available at
\url{http://www.unicode.org/consortium/utc-minutes.html}.}

@@

% ==========================
\section{Pitfall: Encodings}
\label{encodings}
% ==========================

@@

\begin{itemize}
   \item To prevent strange boxes instead of nice glyphs, always install a few
         fonts with a large glyph collection and at least one fall-back font
         (see Section~\ref{pitfall-missing-glyphs}).
   \item Unexpected visual impressions of symbols does not necessarily mean that
         the actual encoding is wrong. It is mostly a problem of faulty
         rendering (see Section~\ref{pitfall-faulty-rendering}).
   \item Do not trust the names of code points as a definition of the character
         (see Section~\ref{pitfall-names}). Also do not trust Unicode blocks as
         a strategy to find specific characters (see
         Section~\ref{pitfall-blocks})
   \item To ensure consistent encoding of texts, apply Unicode normalization
         (NFC or NFD, see Section~\ref{pitfall-canonical-equivalence}).
   \item To prevent remaining inconsistencies after normalization, for example 
         stemming from homoglyphs (see Section~\ref{pitfall-homoglyphs}) 
         or from missing canonical equivalence in the Unicode Standard
         (see Section~\ref{pitfall-absence-of-equivalence})
         use orthography profiles (see Chapter~\ref{orthography-profiles}).
   \item To deal with tailored grapheme clusters
         (Section~\ref{pitfall-characters-are-not-graphemes}), use Unicode Locale 
         Descriptions, or orthography profiles 
         (see Chapter~\ref{orthography-profiles}).
   \item As a preferred file format, use Unicode Format UTF-8 in 
         Normalization Form Composition (NFC) with LF line endings, 
         but without byte order mark (BOM), whenever possible (see 
         Section~\ref{pitfall-file-formats}). This last nicely cryptic 
         recommendation has T-shirt potential:
